Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging

### 2012-09-11 HRUIonGPU/conf_article.tex
started on IUS article

%% bare_jrnl.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% see http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

@@

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[journal]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

@@

%\ifCLASSOPTIONcaptionsoff
%  \usepackage[nomarkers]{endfloat}
% \let\MYoriglatexcaption\caption
% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\fi
% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
% This package may be useful when used in conjunction with IEEEtran.cls'
% captionsoff option. Some IEEE journals/societies require that submissions
% have lists of figures/tables at the end of the paper and that
% figures/tables without any captions are placed on a page by themselves at
% the end of the document. If needed, the draftcls IEEEtran class option or
% \CLASSINPUTbaselinestretch interface can be used to increase the line
% spacing as well. Be sure and use the nomarkers option of endfloat to
% prevent endfloat from "marking" where the figures would have been placed
% in the text. The two hack lines of code above are a slight modification of
% that suggested by in the endfloat docs (section 8.3.1) to ensure that
% the full captions always appear in the list of figures/tables - even if
% the user used the short optional argument of \caption[]{}.
% IEEE papers do not typically make use of \caption[]'s optional argument,
% so this should not be an issue. A similar trick can be used to disable
% captions of packages such as subfig.sty that lack options to turn off
% the subcaptions:
% For subfig.sty:
%\let\MYorigsubfloat\subfloat
%\renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
% For subfigure.sty:
% \let\MYorigsubfigure\subfigure
% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
% However, the above trick will not work if both optional arguments of
% the \subfloat/subfig command are used. Furthermore, there needs to be a
% description of each subfigure *somewhere* and endfloat does not add
% subfigure captions to its list of figures. Thus, the best approach is to
% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
% and instead reference/explain all the subfigures within the main caption.
% The latest version of endfloat.sty and its documentation can obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
%
% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
% later in the document, say, to conditionally put the References on a 
% page by themselves.

@@

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{J. P. \AA{}sen, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
%\thanks{M. Shell is with the Department
%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, January~2013?}%
{\AA{}sen \MakeLowercase{\textit{et al.}}: Capon Adaptive Beamforming on the GPU}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

@@

\begin{abstract}
%\boldmath
An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Adaptive Beamforming, Capon Beamformer, Minimum Variance Beamforming, Graphics Processing Unit, GPGPU, Parallel Computing, Ultrasound Imaging.
\end{IEEEkeywords}

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.

%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
\IEEEPARstart{C}{apon} adaptive beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}. It has been shown that the method introduces some interesting trade-offs in addition to just improved resolution, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}. Despite these benefits, widespread use has been limited by the immense computational complexity associated with the method \cite{So2011}.  In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting. To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows.

Previous work on implementing a parallel Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
%A systolic array processor is custom made hardware, and even if the computational throughput might be higher for selected problems compared with a GPU, the cost related to development and flexibility is considered much higher. 
For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011} have implemented a Capon beamformer on the GPU that produced real apodization weights based on real data. Under this constraint the Capon beamformer is not allowed to shift the main lobe, and therefore its full potential is not reached. Shifting the main lobe is an important property of the Capon beamformer in order for it to improve edge definitions and contrast throughout an ultrasound image. Another aspect of using complex data is that even though the number of aritmetic operations per sample is increased we can now sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum. All results in this paper is reported with the use of baseband data.

%Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced using real data compared with use of complex data. It is crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows. All results in this paper is reported with use of critically sampled complex data, both laterally and in range.  (ikke helt kritisk samplet, ikke si det, bare at det er i baseband).

% denne må utelates i IUS-artikkelen
In recent work (cite JOE) we have applied a GPU-based Capon beamformer in an active sonar system. For sonar the real-time requirement is much lower than for medical ultrasound imaging, which means that most common array configurations can be handled using the GPU without taking any additional measures. For medical ultrasound imaging, and especially cardiac ultrasound imaging, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of Capon beamforming. Since the execution time for a Capon beamformer depends highly on the number of elements in the given array, it is important to find ways of reducing the computational complexity for high-channel-count systems. Different approaches to low-complexity Capon beamforming have recently been proposed \cite{Asl2012, Jensen2012}, and in this paper we have selected to implement the beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009} on the GPU to maintain high framerates as the channel count increases.  
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

In the next section we present background information about the Capon beamformer and the GPU computing model. In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:bs} we introduce the method applied when the problem size increases. Benchmark results and resulting images are presented in Section \ref{sec:res}, and finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

\section{Background}

%\subsection{Capon Beamforming}

%\textbf{Bruk riktig terminologi. $N_{avg}$ er K. Antall sub-arrayer må hete noe annet. F.eks $N_L$}

%\textbf{Presis på hvilke formler vi skal implementere.}

%\textbf{De skal forstå MV etter denne seksjonen.}

%\textbf{Ton ned introduksjonen, og sett opp akkurat de ligningene vi implementerer.}

%\textbf{(Skriv om den introduserende setningen, start rett på linje to, Kombinere data fra mange sensorer, Leseren kan Ultralyd ikke gpu)}

The conventional form of array beamforming is delay-and-sum (DAS). Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ from time $n = 0$ to $N-1$ is delayed to broadside with an appropriate delay $\Delta_m[n]$. The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}^{\Delta}[n]. \label{eq:z}
\end{align}
In conventional beamforming, the weighting $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).

@@

The Capon weight vector is given by the following expression:
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} = \frac{\vec{b}}{\vec{a}^H\vec{b}},
\end{align}
where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}_M$ when $\vec{x}$ is pre-delayed, $\mat{\hat{R}}$ is the sample covariance matrix calculated from the data vector $\vec{x}^\Delta[n]$, and $\vec{b} = \mat{\hat{R}}[n]^{-1}\vec{a}$. 
%\begin{align}
%\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H.
%\end{align} 
%Typically the sample covariance matrix has to be averaged over several samples in order to ensure numerical stability and statistical robustness.

In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over a long sequence of samples\cite{Krima}. For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited. The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned. Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}. In this way, the beamformer output can be too low at broadside, even when there is signal present and the distortionless-response criterion is applied.

In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}. Where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period. We therefore calculate
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n']\vec{x}_l^{\Delta}[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of samples included in time, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$. The dimension of the covariance matrix after this smoothing is $L \times L$. Hence, the size of $\vec{w}[n]$ and $\vec{a}$ in (\ref{eq:w}) has to be $L \times 1$ when the covariance matrix in (\ref{eq:R}) is used.
Finally, the covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability, 
\begin{align}\label{eq:diag}
\mat{\hat{R}} =  f(\vec{x}^{\Delta}[n]) = \mat{\breve{R}} + \epsilon\mat{I}.
\end{align}
A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon &= d \times tr\{\mat{\breve{R}}\}, \\
d &= 1/L
\end{align}
introduced by Synnev\aa{}g et al. \cite{Synnevag2007} has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Nilsen2009, Wang2009, Mehdizadeh2012}. This form of adaptive diagonal loading has also been applied to HF and VHF antenna processing \cite{Featherstone1997}.

The beamformer output is in the end calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights.
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n] \label{eq:z_mv}\\
&= 1/N_L \, \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}^{\Delta}[n] \label{eq:z_mv2},
\end{align}
%where (this matrix is wrong)
%\begin{align}
%\mat{C} = \left(
%\begin{matrix}
%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
%\end{matrix}
%\right)^T
%\end{align}
where $\text{CONV}_m$ is a non-truncated convolution across elements.
Note how we can either select to sum all subarrays first in (\ref{eq:z_mv}), or accumulate the right weighting per element in (\ref{eq:z_mv2}).

% Få med at vi processerer hele bildet på en gang. Forklar forskjellen med å prosessere en beam av gangen. Hva med MLA?
% Illustrer tråder og tider for hvordan processeringen går.
% Vi snakker om mottak-stråleformeren.

%\begin{figure}
%\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}
%\caption{}
%\label{fig:mv}
%\end{figure}

%Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change. 
%In this paper we introduce the first GPU implementation of the Capon beamformer capable of processing a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing. Both types of smoothing are important for maintaining delay-and-sum-like speckle and at the same time keep the high resolution property. This is achieved using complex base band data, which reduces the size of each range line to a minimum. In addition, complex data results in complex weights; hence this is a correct implementation of Capon beamforming with all degrees of freedom preserved. For high channel count systems we propose two ways of reducing computational complexity while maintaining

\subsection{GPU Compute Model}

%\textbf{(Ta med mer banale ting, dette er GPU for ultralydfolk)}

%\textbf{(Etter denne så må de forstå GPU og hvorfor MV passer.)}

In this section we briefly discuss the principle behind GPU computing, laying the basis for further discussion on parallel implementation strategies. CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}. However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper apply for other massive parallel implementation environments as well, like e.g. OpenCL. The key point is that GPUs can accelerate small concurrent tasks, like beamforming one pixel, much better than a CPU. Where the CPU has few cores and logic to process complex threads, the GPU has many cores, but each thread needs to be simple in order to benefit from the extra computational power.

The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function describe the work to be done by each thread. As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be synchronized with little cost if needed. 
%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
From a CUDA perspective the GPU consist of one or more stream multi processors (SMs), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions. A group of blocks is scheduled to each SM until all blocks have been processed.

@@

%When implementing algorithms on the GPU it is important to keep data as close to the core as possible. Utilizing available registers and shared memory to its maximum is therefore important.

>>>>

%% bare_conf.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

@@

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[conference]{IEEEtran}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

@@

@@

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{
\IEEEauthorblockN{
Jon Petter \AA{}sen\IEEEauthorrefmark{1},
Jo Inge Buskenes\IEEEauthorrefmark{2},
Carl-Inge C. Nilsen\IEEEauthorrefmark{2},
Andreas Austeng\IEEEauthorrefmark{2},
Sverre Holm\IEEEauthorrefmark{1}\IEEEauthorrefmark{2}
}
\IEEEauthorblockA{
\IEEEauthorrefmark{1} Mi Lab, Norwegian University of Science and Technology, Trondheim, Norway,\\
\IEEEauthorrefmark{2} Department of Informatics, University of Oslo, Oslo, Norway
}
}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
% 
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3}, 
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

@@

\begin{abstract}
\boldmath
Recent results \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012} suggest that the Capon beamformer can provide increased lateral resolution to a medical ultrasound image. In this paper, the high computational complexity of the Capon beamformer is targeted with the use of Graphics Processing Units (GPUs), and we show in-vivio results with Capon beamforming applied on a full cardiac cycle in addition to simulations. With GPUs we are able to process a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.

For a typical cardiac ultrasound image of $80\times400$ pixels (70 degree sector, 15cm range) acquired using a 2.5MHz, M=64 element phased array, we obtain 10fps (subarray length L=M/2, temporal smoothing over 3 samples). If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44fps. For a 32 element phased array we need less beams to cover the sector ($40\times400$ pixels), hence with the same parameters the frame rate increases to 87fps. The bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels. This is an $O(L^3)$ operation, and results show that when L is larger than 16 (M = 64), real time performance is not possible for cardiac imaging with current high-end hardware. The target GPU has been Nvidia’s Quadro 6000, capable of 1Tflops.

\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

% no keywords

@@

\section{Introduction}
Capon adaptive beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}. For instance it has been introduced some interesting trade-offs in addition to just improved resolution, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}. Despite these benefits, widespread use has been limited by the immense computational complexity associated with the method \cite{So2011}.  In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting. To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPUs), combined with novel methods to reduce the computational complexity as the problem size grows.

Previous work on implementing a parallel Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Sinha2002}. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011} have implemented a Capon beamformer on the GPU that produced real apodization weights based on real data. Under this constraint the Capon beamformer is not allowed to shift the main lobe, and therefore its full potential is not reached. Shifting the main lobe is an important property of the Capon beamformer in order for it to improve edge definitions and contrast throughout an ultrasound image. Another aspect of using complex data is that even though the number of aritmetic operations per sample is increased we can now sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum. All results in this paper is reported with the use of baseband data.

%In the next section we present background information about the Capon beamformer and the GPU computing model. In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:bs} we introduce the method applied when the problem size increases. Benchmark results and resulting images are presented in Section \ref{sec:res}, and finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

\section{Background}
\subsection{Capon Adaptive Beamforming}
The conventional form of array beamforming is delay-and-sum (DAS). Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ from time $n = 0$ to $N-1$ is delayed to broadside with an appropriate delay $\Delta_m[n]$. The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}^{\Delta}[n]. \label{eq:z}
\end{align}
In conventional beamforming, the weighting $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).

@@

To calculate the Capon beamformer output we need three equations. First the sample covariance matrix has to be estimated,
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n']\vec{x}_l^{\Delta}[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of samples included in time, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$. Note that the dimension of the covariance matrix is $L \times L$. The covariance matrix is also loaded with a diagonal factor $\epsilon$ to ensure numerical stability, 
\begin{align}\label{eq:diag}
\mat{\hat{R}} = \mat{\breve{R}} + \epsilon\mat{I}.
\end{align}
A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon &= d \times tr\{\mat{\breve{R}}\}, \\
d &= 1/L
\end{align}
has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Mehdizadeh2012}.

Second, we calculate the Capon weight vector as
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} = \frac{\vec{b}}{\vec{a}^H\vec{b}},
\end{align}
where $\vec{a} = [1_0, 1_1, ..., 1_{L-1}]^T = \vec{1}_L$ when $\vec{x}$ is pre-delayed.

And last, the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights,
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n]. \label{eq:z_mv}
\end{align}

%The Capon weight vector is given by the following expression:
%\begin{align}\label{eq:w}
%\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} = \frac{\vec{b}}{\vec{a}^H\vec{b}},
%\end{align}
%where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}_M$ when $\vec{x}$ is pre-delayed, $\mat{\hat{R}}$ is the sample covariance matrix calculated from %the data vector $\vec{x}^\Delta[n]$, and $\vec{b} = \mat{\hat{R}}[n]^{-1}\vec{a}$.

%In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}. Where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period. We therefore calculate
%\begin{align}
%\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n']\vec{x}_l^{\Delta}[n']^H,\label{eq:R}
%\end{align}
%where  $N_K = 2K + 1$ is the number of samples included in time, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$. The dimension of the covariance matrix after this smoothing is $L \times L$. Hence, the size of $\vec{w}[n]$ and $\vec{a}$ in (\ref{eq:w}) has to be $L \times 1$ when the covariance matrix in (\ref{eq:R}) is used.
%Finally, the covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability, 
%\begin{align}\label{eq:diag}
%\mat{\hat{R}} = \mat{\breve{R}} + \epsilon\mat{I}.
%\end{align}
%A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\breve{R}}$, 
%\begin{align}\label{eq:diag_adapt}
%\epsilon &= d \times tr\{\mat{\breve{R}}\}, \\
%d &= 1/L
%\end{align}
%has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Wang2009, Mehdizadeh2012}.

%The beamformer output is in the end calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights,
%\begin{align}
%z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n]. \label{eq:z_mv}
%\end{align}

\subsection{GPU Compute Model}
CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}. The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function describe the work to be done by each thread. As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be synchronized with little cost if needed. 
%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
From a CUDA perspective the GPU consist of one or more stream multi processors (SMs), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions. A group of blocks is scheduled to each SM until all blocks have been processed.

@@

\section{Parallel Capon}\label{sec:meth}
If we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be further divided into a small set of parallel processes. That is, for each data vector $\vec{x}^\Delta$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
\text{2) } &\text{Solve the linear system of equations } \vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
%\end{array}
\end{align*}

The computational complexity of these three steps is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively.
